---
title: 'Weekly Exercises #4'
author: "Cecelia Kaufmann"
output: 
  html_document:
    keep_md: TRUE
    toc: TRUE
    toc_float: TRUE
    df_print: paged
    code_download: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error=TRUE, message=FALSE, warning=FALSE)
```

```{r libraries}
library(tidyverse)     # for data cleaning and plotting
library(lubridate)     # for date manipulation
library(openintro)     # for the abbr2state() function
library(palmerpenguins)# for Palmer penguin data
library(maps)          # for map data
library(ggmap)         # for mapping points on maps
library(gplots)        # for col2hex() function
library(RColorBrewer)  # for color palettes
library(sf)            # for working with spatial data
library(leaflet)       # for highly customizable mapping
library(carData)       # for Minneapolis police stops data
library(ggthemes)      # for more themes (including theme_map())
theme_set(theme_minimal())
```

```{r data}
# Starbucks locations
Starbucks <- read_csv("https://www.macalester.edu/~ajohns24/Data/Starbucks.csv")

starbucks_us_by_state <- Starbucks %>% 
  filter(Country == "US") %>% 
  count(`State/Province`) %>% 
  mutate(state_name = str_to_lower(abbr2state(`State/Province`))) 

# Lisa's favorite St. Paul places - example for you to create your own data
favorite_stp_by_lisa <- tibble(
  place = c("Home", "Macalester College", "Adams Spanish Immersion", 
            "Spirit Gymnastics", "Bama & Bapa", "Now Bikes",
            "Dance Spectrum", "Pizza Luce", "Brunson's"),
  long = c(-93.1405743, -93.1712321, -93.1451796, 
           -93.1650563, -93.1542883, -93.1696608, 
           -93.1393172, -93.1524256, -93.0753863),
  lat = c(44.950576, 44.9378965, 44.9237914,
          44.9654609, 44.9295072, 44.9436813, 
          44.9399922, 44.9468848, 44.9700727)
  )

#COVID-19 data from the New York Times
covid19 <- read_csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv")

```

## Instructions

* Put your name at the top of the document. 

* **For ALL graphs, you should include appropriate labels.** 

* Feel free to change the default theme, which I currently have set to `theme_minimal()`. 

* Use good coding practice. Read the short sections on good code with [pipes](https://style.tidyverse.org/pipes.html) and [ggplot2](https://style.tidyverse.org/ggplot2.html). **This is part of your grade!**

* When you are finished with ALL the exercises, uncomment the options at the top so your document looks nicer. Don't do it before then, or else you might miss some important warnings and messages.


## Warm-up exercises from tutorial

These exercises will reiterate what you learned in the "Mapping data with R" tutorial. If you haven't gone through the tutorial yet, you should do that first.

### Starbucks locations (`ggmap`)

  1. Add the `Starbucks` locations to a world map. Add an aesthetic to the world map that sets the color of the points according to the ownership type. What, if anything, can you deduce from this visualization?  
  
  I feel the biggest thing I can deduce from such broad data is that many of the Starbucks locations on the East and West Coast are Company Owned, whereas many in the Midwest seem to be Licensed. Many to all locations in Japan are Joint-Ventures. 
  

```{r, message=FALSE}
world <- get_stamenmap(
    bbox = c(left = -180, 
             bottom = -57, 
             right = 179, 
             top = 82.1), 
    maptype = "terrain",
    zoom = 2)

# Plot the points on the map
ggmap(world) + # creates the map "background"
  geom_point(data = Starbucks, 
             aes(x = Longitude, 
                 y = Latitude,
                 color = `Ownership Type`), 
             alpha = .5, 
             size = .3) +
  theme_map() +
  theme(legend.background = element_blank(),
        legend.position = "bottom")

```


  2. Construct a new map of Starbucks locations in the Twin Cities metro area (approximately the 5 county metro area). 
  
  Removing the messages on #2 because the reason some data is being removed is the "zoom" of the map (just the TC). 
  
```{r, message=FALSE}
twincities <- get_stamenmap(
    bbox = c(left = -93.6543, 
             bottom = 44.7266, 
             right = -92.7452, 
             top = 45.2018), 
    maptype = "terrain",
    zoom = 10)

ggmap(twincities) +
  geom_point(data = Starbucks,
             aes(x = Longitude,
                 y = Latitude),
             alpha = .6,
             size = 1.5) +
  theme_map()
```
  

  3. In the Twin Cities plot, play with the zoom number. What does it do?  (just describe what it does - don't actually include more than one map). 

When playing with the zoom number, it seems that the map itself (or the terrain underneath the points plotted), becomes distorted.


  4. Try a couple different map types (see `get_stamenmap()` in help and look at `maptype`). Include a map with one of the other map types.  
```{r, message=FALSE}
twincities <- get_stamenmap(
    bbox = c(left = -93.6543, 
             bottom = 44.7266, 
             right = -92.7452, 
             top = 45.2018), 
    maptype = "toner-lite",
    zoom = 10)

ggmap(twincities) +
  geom_point(data = Starbucks,
             aes(x = Longitude,
                 y = Latitude),
             alpha = .6,
             size = 1.5) +
  theme_map()
```
  

  5. Add a point to the map that indicates Macalester College and label it appropriately. There are many ways you can do think, but I think it's easiest with the `annotate()` function (see `ggplot2` cheatsheet).
  
```{r}
twincities <- get_stamenmap(
    bbox = c(left = -93.6543, 
             bottom = 44.7266, 
             right = -92.7452, 
             top = 45.2018), 
    maptype = "terrain",
    zoom = 10)

ggmap(twincities) +
  geom_point(data = Starbucks,
             aes(x = Longitude,
                 y = Latitude),
             alpha = .6,
             size = 1.5) +
  annotate(geom = "point", 
           x = -93.1691, 
           y = 44.9379, 
           color = "blue") +
  annotate(geom = "text", 
           x = -93.1691, 
           y =44.9379, 
           label = "Macalester College", 
           size = 3) +
  theme_map()

```
  

### Choropleth maps with Starbucks data (`geom_map()`)

The example I showed in the tutorial did not account for population of each state in the map. In the code below, a new variable is created, `starbucks_per_10000`, that gives the number of Starbucks per 10,000 people. It is in the `starbucks_with_2018_pop_est` dataset.

```{r}
census_pop_est_2018 <- read_csv("https://www.dropbox.com/s/6txwv3b4ng7pepe/us_census_2018_state_pop_est.csv?dl=1") %>% 
  separate(state, into = c("dot","state"), extra = "merge") %>% 
  select(-dot) %>% 
  mutate(state = str_to_lower(state))

starbucks_with_2018_pop_est <-
  starbucks_us_by_state %>% 
  left_join(census_pop_est_2018,
            by = c("state_name" = "state")) %>% 
  mutate(starbucks_per_10000 = (n/est_pop_2018)*10000)
```

  6. **`dplyr` review**: Look through the code above and describe what each line of code does.
  
132: creates data set called "census_pop_est_2018" from the dropbox website. 
133: separates state column into variables called "dot" and "state" and then "merge" splints at most (2) times. 
134: select the "dot" variable
135: create a new variable called "state" that converts the state names to lowercase
137 and 138: new data set from starbucks_us_by_state called tarbucks_with_2018_pop_est. 
139: left join the starbucks_us_by_state data with census_pop_est_2018 table 
140: by the variables titles "state_name" and "state"
141: mutate to create a new variable titles "starbucks_per_1000" which is number over "est_pop_2018"/10000. 


  7. Create a choropleth map that shows the number of Starbucks per 10,000 people on a map of the US. Use a new fill color, add points for all Starbucks in the US (except Hawaii and Alaska), add an informative title for the plot, and include a caption that says who created the plot (you!). Make a conclusion about what you observe.
  
  For the most part, the West Coast seems to have the most amount of Starbucks per 10,000 people. 

```{r}
states_map <- map_data("state")

usa_wo_ALHI <- Starbucks %>%
  filter(`Country`== "US",
         !`State/Province`%in% c("AK", "HI"))

# map that colors state by number of Starbucks
starbucks_with_2018_pop_est %>% 
  ggplot() +
  geom_map(map = states_map,
           aes(map_id = state_name,
               fill = starbucks_per_10000)) +
  geom_point(data = usa_wo_ALHI,
             aes(x = Longitude,
                 y = Latitude)) +
  scale_fill_viridis_c() +
  #This assures the map looks decently nice:
  expand_limits(x = states_map$long, 
                y = states_map$lat) + 
  theme_map() +
  theme(legend.background = element_blank(),
        legend.position = "bottom") +
  labs(fill = "Starbucks per 10000 people",
       title = "Proportions of Starbucks and 
       Location of Starbucks in the US Show the Most Starbucks are on the East and West Coast",
       subtitle = "We can see that the West Coast has both the many Stabucks locations as well as the highest concentration per 10,000 people.",
       caption = "Graph Created by: Cecelia Kaufmann")
```
  

### A few of your favorite things (`leaflet`)

  8. In this exercise, you are going to create a single map of some of your favorite places! The end result will be one map that satisfies the criteria below. 

  * Create a data set using the `tibble()` function that has 10-15 rows of your favorite places. The columns will be the name of the location, the latitude, the longitude, and a column that indicates if it is in your top 3 favorite locations or not. For an example of how to use `tibble()`, look at the `favorite_stp_by_lisa` I created in the data R code chunk at the beginning.  
```{r}
favorite_tc_by_cecelia <- tibble(
  place = c("Milkweed Coffee", "Sift Gluten-Free Bakery", "The Guthrie", 
            "Minehaha Falls", "Hidden Falls", "South High School",
            "Electric Fetus", "United Noodles", "Whole Foods", "Pho 79"),
  long = c(-93.2168, -93.2521, -93.2559, 
           -93.2110, -93.1543, -93.2446, 
           -93.2704, -93.2421,-93.1667, 
           -93.1950),
  lat = c(44.9486, 44.9199, 44.9782,
          44.9153, 44.90968, 44.9455,
          44.96259, 44.9583, 44.9471, 
          44.9748), 
  top_three = c("Yes", "No", "No", 
  "No", "Yes", "No", 
  "No", "No", "Yes", "No")
  
  )
favorite_tc_by_cecelia

```
  

  * Create a `leaflet` map that uses circles to indicate your favorite places. Label them with the name of the place. Choose the base map you like best. Color your 3 favorite places differently than the ones that are not in your top 3 (HINT: `colorFactor()`). Add a legend that explains what the colors mean.  
  
  P.S. As you can see, there is a legend in the code... and it's showing up in my R studio right now as I type this, I do not know why it is not showing up in the kitted file..., this is also the case for the next problem as well. 

```{r}
pal_color <- colorFactor("Spectral",
                         domain = favorite_tc_by_cecelia$top_three)

leaflet(data = favorite_tc_by_cecelia) %>% #base plot
  addProviderTiles(providers$CartoDB.DarkMatter) %>% 
  addCircles(lng = ~long, 
             lat = ~lat,
             label = ~place,
             opacity = 1,
             radius = 50,
             color = ~pal_color(top_three)) %>%
  addLegend(position = "bottomright",
            pal = pal_color,
            values = ~top_three,
            opacity = 1,
            title = "Top Three") 
```
  
  
  * Connect all your locations together with a line in a meaningful way (you may need to order them differently in the original data).  
  
Honestly, I drew the lines the way I entered the data and it looks really cool and kind of like a star... so I kept it! I hope that is in fact meaningful enough!!


```{r}
pal_color <- colorFactor("Spectral",
                         domain = favorite_tc_by_cecelia$top_three)

leaflet(data = favorite_tc_by_cecelia) %>% #base plot
   addProviderTiles(providers$CartoDB.DarkMatter) %>%
  addCircles(lng = ~long, 
             lat = ~lat,
             label = ~place,
             opacity = 1,
             radius = 50,
             color = ~pal_color(top_three)) %>%
  addLegend(position = "bottomright",
            pal = pal_color,
            values = ~top_three,
            opacity = 1,
            title = "Top Three") %>%
   addPolylines(lng = ~long, 
               lat = ~lat, 
               color = col2hex("darkred"))
```
  
  * If there are other variables you want to add that could enhance your plot, do that now.  
  
## Revisiting old datasets

This section will revisit some datasets we have used previously and bring in a mapping component. 

### Bicycle-Use Patterns

The data come from Washington, DC and cover the last quarter of 2014.

Two data tables are available:

- `Trips` contains records of individual rentals
- `Stations` gives the locations of the bike rental stations

Here is the code to read in the data. We do this a little differently than usualy, which is why it is included here rather than at the top of this file. To avoid repeatedly re-reading the files, start the data import chunk with `{r cache = TRUE}` rather than the usual `{r}`. This code reads in the large dataset right away.

```{r cache=TRUE}
data_site <- 
  "https://www.macalester.edu/~dshuman1/data/112/2014-Q4-Trips-History-Data.rds" 
Trips <- readRDS(gzcon(url(data_site)))
Stations<-read_csv("http://www.macalester.edu/~dshuman1/data/112/DC-Stations.csv")
```

  9. Use the latitude and longitude variables in `Stations` to make a visualization of the total number of departures from each station in the `Trips` data. Use either color or size to show the variation in number of departures. This time, plot the points on top of a map. Use any of the mapping tools you'd like.
  
  I don't know how to change the scale range, or consolidate it so it doesn't go on for a long time. 
  
```{r}
newstationsandtrips <- Stations %>%
  left_join(Trips,
            by = c("name" = "sstation")) %>%
  group_by(long, lat) %>%
    summarise(numtimes = n())


pal_color <- colorFactor("plasma",
                         domain = newstationsandtrips$numtimes)

leaflet(data = newstationsandtrips) %>% #base plot
  addTiles() %>% #base map - default is openstreet map 
  addCircles(lng = ~long, 
             lat = ~lat,
             weight = 5,
             opacity = .7,
             radius = 50,
             color = ~pal_color(numtimes)) %>%
  addLegend(position = "topright",
            pal = pal_color,
            values = ~numtimes,
            opacity = 1,
            title = "Number of Times Visited")

```
  
  10. Only 14.4% of the trips in our data are carried out by casual users. Create a plot that shows which area(s) have stations with a much higher percentage of departures by casual users. What patterns do you notice? Also plot this on top of a map. I think it will be more clear what the patterns are.
  
  We can reasonable assume upon seeing the map that many of the casual riders take these trips around monuments or other types of attractions in the city, like noted on the map (i.e. National Mall). We also see high proportions by the Pentagon and Ft. Barnard, which could be tourist attractions as well as the Smithsonian, which we know is a common tourist attraction in DC. 
  
```{r}
propcasrides <- Stations %>%
  left_join(Trips,
            by = c("name" = "sstation")) %>%
  group_by(long, lat) %>%
  summarise(numtimes = n(),
         propridescas = mean(client == "Casual")) 

pal_color <- colorFactor("plasma",
                         domain = propcasrides$propridescas)

leaflet(data = propcasrides) %>% #base plot
  addTiles() %>% #base map - default is openstreet map 
  addCircles(lng = ~long, 
             lat = ~lat,
             weight = 5,
             opacity = .7,
             radius = 50,
             color = ~pal_color(propridescas)) 

```
  
### COVID-19 data

The following exercises will use the COVID-19 data from the NYT.

  11. Create a map that colors the states by the most recent cumulative number of COVID-19 cases (remember, these data report cumulative numbers so you don't need to compute that). Describe what you see. What is the problem with this map?
  
  I feel like this grossly misrepresents the number of COVID cases, particularly the scale would be difficult to read for the general public (1e +06 etc.). Also, it should California having off the charts numbers, which may be true, but grossly misrepresents the rest of the country when considering the population of California versus say, South Dakota. 
  
  

```{r}
states_map <- map_data("state")
covid19 %>%
  group_by(state) %>%
  summarise(maxcases = max(cases)) %>%
  mutate(state = str_to_lower(state)) %>%
  ggplot(aes(fill = maxcases)) +
  geom_map(map = states_map,
           aes(map_id = state)) +
   scale_fill_distiller(palette = "Spectral") + 
  expand_limits(x = states_map$long, 
                y = states_map$lat) + 
  theme_map() +
  theme(legend.background = element_blank(),
        legend.position = "bottom") +
  labs (title = "Cumulative Number of Covid Cases in Each State",
        fill = "Cumulative Cases")

```
  
  
  12. Now add the population of each state to the dataset and color the states by most recent cumulative cases/10,000 people. See the code for doing this with the Starbucks data. You will need to make some modifications. 
  
  This map shows a MUCH better representation!!!
  
```{r}
covid19 %>%
  group_by(state) %>%
  summarise(maxcases = max(cases)) %>%
  mutate(state = str_to_lower(state)) %>%
  left_join(census_pop_est_2018,
            by = c("state")) %>%
    mutate(covid_per_10000 = (maxcases/est_pop_2018)*10000) %>%
  ggplot(aes(fill = covid_per_10000)) +
  geom_map(map = states_map,
           aes(map_id = state)) +
   scale_fill_distiller(palette = "Spectral") + 
  expand_limits(x = states_map$long, 
                y = states_map$lat) + 
  theme_map() +
  theme(legend.background = element_blank(),
        legend.position = "bottom") +
  labs (title = "Proportion of Cumulative Covid-19 Cases Per 10,000 People",
        fill = "Proportion (or Percent) Per 10,000 People")


```
  
  
  13. **CHALLENGE** Choose 4 dates spread over the time period of the data and create the same map as in exercise 12 for each of the dates. Display the four graphs together using faceting. What do you notice?
```{r}

```
  
  
## Minneapolis police stops

These exercises use the datasets `MplsStops` and `MplsDemo` from the `carData` library. Search for them in Help to find out more information.

  14. Use the `MplsStops` dataset to find out how many stops there were for each neighborhood and the proportion of stops that were for a suspicious vehicle or person. Sort the results from most to least number of stops. Save this as a dataset called `mpls_suspicious` and display the table.  

```{r, message=FALSE}
mpls_suspicious <- MplsStops %>%
   group_by(neighborhood) %>%
  summarise(n = n(),
            prop_sus = mean(problem == "suspicious"))%>%
  arrange(desc(n))
mpls_suspicious
```
  
  
  15. Use a `leaflet` map and the `MplsStops` dataset to display each of the stops on a map as a small point. Color the points differently depending on whether they were for suspicious vehicle/person or a traffic stop (the `problem` variable). HINTS: use `addCircleMarkers`, set `stroke = FAlSE`, use `colorFactor()` to create a palette.  
  
```{r}
pal_color <- colorFactor("plasma",
                         domain = MplsStops$problem)

leaflet(data = MplsStops) %>% #base plot
  addTiles() %>% #base map - default is openstreet map 
  addCircleMarkers(stroke = FALSE,
                   lng = ~long, 
             lat = ~lat,
             weight = 5,
             opacity = .7,
             radius = .2,
             color = ~pal_color(problem))

```
  
  
  16. Save the folder from moodle called Minneapolis_Neighborhoods into your project/repository folder for this assignment. Make sure the folder is called Minneapolis_Neighborhoods. Use the code below to read in the data and make sure to **delete the `eval=FALSE`**. Although it looks like it only links to the .sph file, you need the entire folder of files to create the `mpls_nbhd` data set. These data contain information about the geometries of the Minneapolis neighborhoods. Using the `mpls_nbhd` dataset as the base file, join the `mpls_suspicious` and `MplsDemo` datasets to it by neighborhood (careful, they are named different things in the different files). Call this new dataset `mpls_all`.

```{r}
mpls_nbhd <- st_read("Minneapolis_Neighborhoods/Minneapolis_Neighborhoods.shp", quiet = TRUE)

mpls_all <- mpls_nbhd %>%
  left_join(MplsDemo,
            by = c("BDNAME"= "neighborhood")) %>%
  left_join(mpls_suspicious,
            by = c("BDNAME"= "neighborhood"))
mpls_all
  
```

  17. Use `leaflet` to create a map from the `mpls_all` data  that colors the neighborhoods by `prop_suspicious`. Display the neighborhood name as you scroll over it. Describe what you observe in the map.
  
  Southeast Minneapolis into Downtown up to basically the I35W line has a much higher rate of suspicious stops versus Southwest and Northeast Minneapolis. Armatage (in SW Mpls), Shingle Creek, and Lind seem to stray from that trend. As a life-long MPLS resident, I can say why that may be the case in the Northern corners of Minneapolis, but Armatage is a little bit of mystery. As an area of MPLS as a whole, NE MPLS seems to have the lowest proportion of suspicious stops overall, which is quite interesting.  
  
```{r}
pal_color <- colorFactor("Reds",
                         domain = mpls_all$prop_sus)

leaflet(data = mpls_all) %>%
  addTiles() %>%
  addPolygons(color = ~pal_color(prop_sus),
              label = ~BDNAME)
  
```
  
  
  18. Use `leaflet` to create a map of your own choosing. Come up with a question you want to try to answer and use the map to help answer that question. Describe what your map shows. 
  
Where are the locations of Starbucks in MN and what ownership style do they have?

We see (and not shockingly) many of the Starbucks in the state are based around the Twin Cities, and furthermore, we can see that only Licensed and Company-Owned stores are in the state of MN. 
  
```{r}
mnStarbucks <- Starbucks %>%
  filter(Country == "US",
         `State/Province`== "MN")

pal_color <- colorFactor("plasma",
                         domain = mnStarbucks$`Ownership Type`)

leaflet(data = mnStarbucks) %>% #base plot
  addTiles() %>% #base map - default is openstreet map 
  addCircles(lng = ~Longitude, 
             lat = ~Latitude,
             weight = 5,
             opacity = .7,
             radius = 50,
             label = ~`Store Name`,
             color = ~pal_color(`Ownership Type`)) %>%
    addLegend(position = "topright",
            pal = pal_color,
            values = ~`Ownership Type`,
            opacity = 1,
            title = "Ownership Type")


```
  
  
  
## GitHub link

  19. Below, provide a link to your GitHub page with this set of Weekly Exercises. Specifically, if the name of the file is 04_exercises.Rmd, provide a link to the 04_exercises.md file, which is the one that will be most readable on GitHub.


**DID YOU REMEMBER TO UNCOMMENT THE OPTIONS AT THE TOP?**
